\textls[-0]{\thetitle}

### 2023-03-13 metatheory2.tex
Start changing on appendices.

In this section, we start by introducing the complete formulation of the syntax of \TT, 
and then we detail the models and proofs for syntactic translation (that translate away linkages), 
consistency and canonicity.

\subsection{Warm Up: Review Declarative Syntax for MLTT}
We begin with a review of MLTT as \TT can be roughly considered as MLTT with linkages.
For our syntax, we work in an intrinsic-typed setting, so there is no un-typed preterms appearing anywhere. We introduce judgemental equality ($\equiv$)
for derivations (of all kinds of judgements). In other words, we will have judgemental equality between terms, 
contexts and explicits substitutions. 
We will make sure all the introduced models and proofs respecting judgemental equality. In other words, we will consider our syntax as the result of taking quotient by these judgemental equality.

@@

% Copy and Paste Intuition of MLTT from other places to here
\subsection{Declarative Syntax for \TT : an MLTT variant with Linkages}
Compared to MLTT, \TT has 3 more judgements about signatures and linkage transformers. Moreover, the linkage transformer can be considered as a syntactic sugar (or library functions) (we will show why at the end). 
Since it is a syntactic sugar, we will not consider about the linkage transformers during meta-theoretic reasoning.

>>>>

In this section, we lay out the complete version of the syntax of \TT but in the style of \citeauthor{debruijn1964} indices and \emph{explicit substitutions}~\cite{substcalculus,abadi1989subst}:
substitutions~$\gm$ and their applications (e.g., in the form of \sub{T}{\gm}) are part
of the syntax rather than meta-operations.

Other parts of the syntax remained the same as the main text.

We also work in an intrinsic-typed setting, so there is no un-typed preterms appearing anywhere. We introduce judgemental equality ($\equiv$)
for derivations (of all kinds of judgements). In other words, we will have judgemental equality between terms, 
contexts and explicits substitutions.

This style of syntax formulation has been carried out by the recent trend~\cite{altkap2016,coquand2018canonicity,gratzer-sterling-birkedal-2019} dubbed as ``algebraic presentation'' of MLTT.

After the syntax, we detail the models and proofs for syntactic translation (that translate away linkages),  consistency and canonicity.

We will make sure all the introduced models and proofs respecting judgemental equality. In other words, we will consider our syntax as the result of taking quotient by these judgemental equality.

\subsection{Review MLTT in Debruijn Indices and Explicit Subst}

\TT can be roughly considered as MLTT with linkages so we still start with vanilla MLTT first.

@@

The change of the syntax formulation is important to the simplicity of our proof, as we want to handle the proof about substitutions and terms uniformly---thus making substitution as part of the syntax (explicit substitution) will be helpful.

Variables are in the form of \citeauthor{debruijn1964} indices:
\var{n} is the variable bound by the $n$-th closest enclosing binder.
For example, $\lambda x.\,\lambda y.\,x$ is $\lam{\lam{\var{1}}}$.
%
Substitutions are typed with the form 
%\EDJ{It seems to me G |- g : D is a judgement not a rule. Rules usually have two parts (premises + conclusion). Judgment is a standard terminology invented by Martin Lof (https://ncatlab.org/nlab/show/judgment see reference here). I will suggest saying 'Formation rules for substitution lead to a judgement G |- g : D' sth like this. In pen-and-paper, the formation rule (corresponding to QIIT well-typed-ness of judgement) is called presuppose (can be found in Modal MLTT paper) }
$\goodSub{\Gm}{\gm}{\Dl}$.
The idea is that applying~$\gm$ to terms valid in the context~$\Dl$
yields terms valid in~$\Gm$ (\ruleref{tm/sub} and \ruleref{ty/sub}).
%
The two main forms of substitutions are weakening (\ruleref{sub/wk}) and extension (\ruleref{sub/ext}):
\sub{t}{\SubstWeak{n}} introduces $n$ free variables into the context of $t$, and
\sub{t}{\SubstExt{\gm}{t'}} substitutes~$t'$ for~$\var{0}$ in~$t$ and
then applies~$\gm$.
For example, rule \ruleref{tm/snd} states that if $t$ is a dependent pair
that has type $\TySigma{A}{B}$, then $\snd{t}$ has type
$\sub{B}{\SubstExt{\SubstWeak{0}}{\fst{t}}}$,
where $\SubstId$ is the identity substitution (\ruleref{sub/id}).

% Copy and Paste Intuition of MLTT from other places to here
\subsection{Declarative Syntax for \TT : an MLTT variant with Linkages}
Compared to MLTT, \TT has 3 more judgements about signatures and linkage transformers. Moreover, the linkage transformer can be considered as a syntactic sugar (or ``library functions'' in the main text) (we will show why at the end). 
Since it is a syntactic sugar, we will not consider about the linkage transformers during meta-theoretic reasoning.

### 2023-03-13 syn-translate.tex
Start changing on appendices.

This ``subpart'' is the standard MLTT with an
(unconventional) formulation of W-types,
and recursors that use $\Sigma$-types to aggregate handlers.

Basically, we can compile away (the derivation of) judgement $\goodSig{\Gamma}{v}{n}$,
the type $\cL$ and $\cC$, and related terms, by transforming them into
corresponding $\Sigma$-types.

@@

We omit all the equational rules ($\beta,\eta$ and substitution) here.
However, for coherence, when we mutual recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}} $, and
$\denotesT{\cP}$ above, we actually have to prove the substitution laws
$\sub{(\TyTkg{\lsig})}{\gm}\equiv \TyTkg{\sub{\lsig}{\gm}}$ and $\sub{(\Sigr{\Delta}{n})}{\gm}
\equiv \Sigr{\Gamma}{n}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.

>>>>

This ``subpart'' is the standard MLTT with our
(unconventional) formulation of W-types,
and recursors that use $\Sigma$-types to aggregate handlers.

Basically, we can compile away (the derivation of) judgement $\goodSig{\Gamma}{v}{n}$,
the type $\cL$, $\cP$, and related terms, by transforming them into
corresponding $\Sigma$-types.

@@

We omit all the validation of equational rules ($\beta,\eta$ and substitution) here.
However, when we mutual recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}} $, and
$\denotesT{\cP}$ above, we actually have to prove the substitution laws
$\sub{(\TyTkg{\lsig})}{\gm}\equiv \TyTkg{\sub{\lsig}{\gm}}$ and $\sub{(\Sigr{\Delta}{n})}{\gm}
\equiv \Sigr{\Gamma}{n}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.
